---
title: "Homework 5"
format: html
editor: visual
---

## Summarizing Student Data

```{r}
# Load libraries
library(tidyverse)
```

### Task 1: Read in the Data and Modify

Modify the code provided to read in the data from a local file source and combine the two data frames. Use local paths as they do in their code.

```{r}
# Code provided
d1=read.table("student+performance/student/student-mat.csv",sep=";",header=TRUE)
d2=read.table("student+performance/student/student-por.csv",sep=";",header=TRUE)

d3=merge(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
print(nrow(d3)) # 382 students
```

Read in and combine the data using functions from the tidyverse. Use an inner_join() on the variables they used in their code. Do you notice any issues? Make a note of the issue.

```{r}
# Load the data
tbl1 <- read_csv2("student+performance/student/student-mat.csv", 
                  col_names=TRUE)
tbl2 <- read_csv2("student+performance/student/student-por.csv",
                  col_names=TRUE)

# inner join

```
